---
title: "Perfomance estimation from high throughput screenings "
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Cecilia Trivellin  
### Industrial Biotechnology Division, Chalmers University of Technology 2021

## Aim of the script

This script aims to calculate performance values from growth curves and sugar/ethanol concentrations in high throughput screenings.
Growth functions (maximum specific growth rate, lag phase) are calculated from growth curves exported from the growth profiler in GV units vs time.
Yields are calculated from enzymatic essays measuring glucose, ethanol and other sugars. 
Each plate analyzed correspond to a specific strain. In the plate various conditions of growth are present.

Information regarding methodology and materials are reported here:
[insert link bioarchives]


## Data implementation

```{r, results='hide'}
#loading libraries
library(tidyverse)
library(lattice)
library(deSolve)
library(growthrates)
library(ggplot2)
library(ggExtra)
library(naniar)
library(ggpubr)
library(statmod)

#loading data
#summary contains information on the layout of the growth_profiler plate
summary <- readxl::read_excel("~/Data/plate_legend_ACS.xlsx")
#data contains the green values "GV" measured in the growth profiler. Each column corresponds to a well, except the 1st column which is the time in minutes.
#the time column is converted in hours, the last 9 column are removed because they are empty
#data is reorganized to have only 3 columns (well, values, time(h))
data <- readxl::read_excel("~/Data/CEN.PK113-7D/CENPK7D_1.xlsx", sheet = "GV") %>%
  mutate(timeh = `time`/60) %>%
  select(-`time`, -`H4`, -`H5`,-`H6`,-`H7`, -`H8`, -`H9`,-`H10`,-`H11`, -`H12`) %>%
  gather("well", "value", -timeh) 
#the summary of plate and the log of the growth values are merged in the same table, the mean and the standard deviation of the replicates are calculated
strain <- inner_join(data, summary, by="well") %>%
  group_by(condition, timeh) %>%
  mutate(log_value=log(value)) %>%
  mutate(mean=mean(log_value)) %>%
  mutate(stdev=sd(log_value)) %>%
  filter(replicate == 1) 
#statistical test to compare the growth curves in different media with the control condition "Glu_20"
stats <- compareGrowthCurves(strain$condition,as.matrix(strain$mean)) %>%
  filter(Group2 == "Glu_20" | Group1 =="Glu_20") 
#swapping columns to plot the data
tmp <- stats[1:12,1]
stats[1:12,1] <- stats[1:12,2]
stats[1:12,2] <- tmp
#plotting P-values
ggboxplot(stats, x = "Group1", y = "P.Value", colour= "Group2", palette = "jco",
          add = "jitter")
#extracting conditions that are significantly different from Glu_20
stats_results <- stats %>% 
  filter(P.Value <= 0.05)
print(stats_results$Group2)

#create a table with all replicates to plot in log scale 
strain_replicates <- right_join(data, summary, by="well") %>%
  mutate(replicate= as.character(replicate)) %>%
  mutate(log_value=log(value))
```

## Growth curves plots

```{r}
palette <- c("#66c2a5", "#fc8d62", "#8da0cb", "#fb8072", "#a6d854", "#ffd92f")

ggplot(data = strain) +
  geom_ribbon(aes(x= timeh, ymin = log_value-stdev, ymax = log_value+stdev, fill= group, alpha= 0.5))+
  geom_line(aes(y=log_value, x=timeh, colour=group)) +
  scale_fill_manual(values = palette) +
  scale_colour_manual(values = palette) +
  labs(title = "CENPK7D",
       subtitle = "growth curves",
       y = "log_GV", x = "time(h)") + 
  facet_wrap(~condition)  +
  theme(legend.title = element_blank(), 
              text = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                              colour = "grey"), 
              panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                              colour = "grey"))
#plot growth curves to visualize replicates
ggplot() +
  geom_point(data = strain_replicates, size = 0.5, aes(y=log_value, x=timeh, colour=group, shape=replicate)) +
  scale_colour_manual(values = palette) +
  labs(title = "CENPK7D",
       subtitle = "growth curves + replicates",
       y = "log_GV", x = "time(h)") + 
  facet_wrap(~condition)  +
  theme(legend.title = element_blank(), 
        text = element_text(size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "grey"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "grey"))
library(RColorBrewer)
nb.cols <- 29
mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(nb.cols)

ggplot(data = strain) +
    geom_ribbon(aes(x= timeh, ymin = mean-stdev, ymax = mean+stdev, fill= condition, alpha= 0.5))+
    geom_line(aes(y=mean, x=timeh, colour=condition)) +
    scale_fill_manual(values = mycolors) +
    scale_colour_manual(values = mycolors)+
    labs(title = "CENPK7D",
         subtitle = "growth curves",
         y = "logGV", x = "time(h)") +
      facet_wrap(~group)  +
  theme(legend.title = element_blank(), 
        text = element_text(size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "grey"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "grey"))

```
## Calculation of maximum specific growth rate and lag phase 

```{r}
#the function all_splines is used to calculate the µmax 
many_spline_fits <- all_splines(value ~ timeh | group + condition + replicate,
                                data = strain_replicates, spar = 0.7)
par(mfrow = c(6,6))
par(mar=c(1,1,1,1))
plot(many_spline_fits)
#data frame is created with the results coming from all_splines function
results <- results(many_spline_fits)

#calculate lambda (lag phase): the coordinates of the inflection point have been extracted and used to calculate the tangent and subsequently the x value corresponding to the lag phase
i=0
inf_points <- NULL
for (i in 1:87){
  tmp <- many_spline_fits@fits[[i]]@xy 
  inf_points <- rbind(inf_points, tmp)
}
inf_points <- as.data.frame(inf_points)
lambda_mu <- bind_cols(results, inf_points) %>% 
  dplyr::rename(x = V1, y = V2) %>%
  mutate(lag=((log10(y0)-log10(y))/mumax)+x) %>%
  mutate(replicate= as.character(replicate))

#all_splines function will approximate each curve. In the wells with no growth, the values reported are wrongly estimated but R2 values are low therefore can be excluded 
lambda_mu$mumax[lambda_mu$r2< 0.985] <- 0
lambda_mu$lag[lambda_mu$r2<0.985] <-0
lambda_mu <- lambda_mu %>%
  select(-`y0`, -`x`, -`y`)

#visualize the replicates with a boxplot
boxplot(mumax ~ condition, data = lambda_mu, las=2, xlab = "",
        ylab = "µmax (1/h)", main = "µmax_replicates")
title(xlab = "Conditions", line = 6)
boxplot(lag ~ condition, data = lambda_mu, las=2, xlab = "",
        ylab = "lag_phase (h)", main = "lag_replicates")
title(xlab = "Conditions", line = 6)
```

## Calculation of yields and cell dry weight
The data imported are measured with enzymatic assays (sugars and ethanol). Yields and cell dry weight (end of cultivation) are calculated. 

```{r}
#load data
strain_summary <- readxl::read_excel("~/Data/CEN.PK113-7D/CENPK7D_summary.xlsx")
summary_dilutions <- readxl::read_excel("~/Data/summary_dilutions.xlsx")
sugar_initial_concentration <- readxl::read_excel("~/Data/sugar_initial_concentration.xlsx")
```

## Calculation of final biomass dry weight
Calculating the cell dry weight with calibration curves previously determined. Slope coefficient converts final OD values in final biomass (g/L). 

```{r}
CDW <- select(strain_summary, well, condition, replicate, group, final_OD, CDW_OD) %>%
  mutate(CDW=final_OD*CDW_OD[1]) %>%
  dplyr::slice(1:87)
```

## Calculation of sugars concentrations

```{r, results='hide'}
#calculate glucose concentrations, postion 89 is the standard for the sugar essays
Glu_conc <- select(strain_summary, well, condition, replicate, group, A1_glu, A0_glu)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`man`, -`gal`, -`xyl`, -`ara`) %>%
  mutate(final_glu_conc=((A1_glu-A0_glu)/(A1_glu[89]-A0_glu[89]))*glu*0.4) %>%
  dplyr::slice(1:87) #to remove the blank and the standards 

#calculte other sugars concentrations 
Man_conc <- select(strain_summary, well, condition, replicate, group, A1_man, A0_man)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`xyl`, -`ara`) %>%
  mutate(final_man_conc=((A1_man-A0_man)/(A1_man[32]-A0_man[32]))*man*0.1) %>%
  dplyr::slice(25:30) 
Xyl_conc <- select(strain_summary, well, condition, replicate, group, A1_xyl, A0_xyl)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`man`, -`ara`) %>%
  mutate(final_xyl_conc=((A1_xyl-A0_xyl)/(A1_xyl[14]-A0_xyl[14]))*xyl*0.25) %>%
  dplyr::slice(7:12)
Gal_conc <- select(strain_summary, well, condition, replicate, group, A1_gal, A0_gal)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`man`, -`xyl`, -`ara`) %>%
  mutate(final_gal_conc=((A1_gal-A0_gal)/(A1_gal[20]-A0_gal[20]))*gal*0.4) %>%
  dplyr::slice(13:18)
Ara_conc <- select(strain_summary, well, condition, replicate, group, A1_ara, A0_ara)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`xyl`, -`man`) %>%
  mutate(final_ara_conc=((A1_ara-A0_ara)/(A1_ara[26]-A0_ara[26]))*ara*0.4) %>%
  dplyr::slice(19:24)
```

## Calculation of biomass and product yields (g/g)

```{r, results='hide'}
#merging tables with sugars, ethanol and biomass concentrations
yields <- select(strain_summary, well, condition, replicate, group, etoh_conc) %>%
  right_join(CDW, by=c("well", "condition", "replicate", "group")) %>%
  select(-`final_OD`, -`CDW_OD`) %>%
  right_join(Glu_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_glu`, -`A0_glu`, -`glu`)%>%
  left_join(Man_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_man`, -`A0_man`, -`man`) %>%
  left_join(Xyl_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_xyl`, -`A0_xyl`, -`xyl`) %>%
  left_join(Gal_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_gal`, -`A0_gal`, -`gal`) %>%
  left_join(Ara_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_ara`, -`A0_ara`, -`ara`) %>%
  replace_with_na_all(condition = ~.x <= 0) %>% #replacing negative value with 0 (in the test it means there is no more sugar to measure)
  replace_na(list(final_glu_conc=0, final_man_conc = 0, final_xyl_conc = 0, final_gal_conc = 0, final_ara_conc = 0 )) %>% #replacing NA values with 0 otherwise it cannot calculate the sum
  left_join(sugar_initial_concentration, by=c("well", "condition", "replicate", "group"))

#calculating yields
 yields <- yields %>% mutate(final_sugars_conc=final_man_conc+final_xyl_conc+final_gal_conc+final_ara_conc) %>%
  select(-`final_man_conc`,-`final_gal_conc`,-`final_xyl_conc`,-`final_ara_conc`)%>%
  dplyr::mutate(consumed_glucose=glu_conc-final_glu_conc) %>%
  select(-`glu_conc`,-`final_glu_conc`) %>%
  dplyr::mutate(consumed_sugarx=other_sugar_conc-final_sugars_conc) %>%
  select(-`other_sugar_conc`,-`final_sugars_conc`) %>%  #calculating how much sugar has been consumed
  replace_with_na_all(condition = ~.x <= 0)%>%  #replacing negative values with 0 
  replace_na(list(consumed_glucose=0, consumed_sugarx = 0, etoh_conc=0)) %>%
  dplyr::mutate(Yx=CDW/(consumed_glucose+consumed_sugarx)) %>% 
  dplyr::mutate(Yp=etoh_conc/(consumed_glucose+consumed_sugarx)) %>%
  select(-`etoh_conc`, -`consumed_glucose`,-`consumed_sugarx`)
  
#replacing infinite values with 0 and ethanol yield values, in media with ethanol, to 0 (enzymatic assays not able to correctly quantify the amount of ethanol due to evaporation and high concentrations)
yields$Yx[yields$Yx == Inf] <-0
yields$Yp[yields$Yp == Inf] <-0
yields$Yp[yields$condition == "EtOH_45"] <-0
yields$Yp[yields$condition == "EtOH_90"] <-0
```

Growth data and yields are merged and visualized

```{r}
#putting together the growth variables and the production variables
lambda_mu$replicate <- as.numeric(as.character(lambda_mu$replicate))

CENPK7D_data_exp <- yields %>%
  inner_join(lambda_mu, by=c("condition", "replicate","group")) %>%
  select(-`well`, -`r2`) %>%
  mutate(strain = "CENPK7D") %>%
  mutate(strain_type ="lab") 

CENPK7D_data_exp$CDW[CENPK7D_data_exp$mumax == 0] <-0
CENPK7D_data_exp$Yx[CENPK7D_data_exp$mumax == 0] <-0
CENPK7D_data_exp$Yp[CENPK7D_data_exp$mumax == 0] <-0

CENPK7D_data <- CENPK7D_data_exp %>%
  group_by(condition) %>%
  mutate(mumax_m=mean(mumax)) %>%
  mutate(lag_m=mean(lag)) %>%
  mutate(CDW_m=mean(CDW)) %>%
  mutate(Yx_m=mean(Yx)) %>%
  mutate(Yp_m=mean(Yp)) %>%
  filter(replicate==1)
  
#reshaping and plotting
CENPK7D_reshaped <- gather(CENPK7D_data, key="variable", value="value", c("mumax_m", "lag_m", "Yx_m", "Yp_m", "CDW_m"))

ggplot(CENPK7D_reshaped, aes(x=condition, y=value, fill=group))+
  scale_fill_manual(values=palette)+
  geom_bar(stat='identity')+
  facet_wrap(~variable,  ncol=1, strip.position = "left", scales="free_y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), text = element_text(size = 10))

```


