---
title: "Robustness quantification"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo= TRUE, warning = FALSE, message = FALSE) 
```

## Cecilia Trivellin  
### Industrial Biotechnology Division, Chalmers University of Technology 2021

## Aim of the script
The following script aim to calculate robustness of cellular functions of three different strains. Cellular functions correspond to: maximum specific growth rate, cell dry weight, biomass and ethanol yield and lag phase.

Information regarding methodology and materials are reported here:
[insert link bioarchives]

## Loading and tidying data

```{r, results='hide'}
#loading libraries
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggsci)
library(RColorBrewer) 
library(ggridges)
library(stats)

#importing performance data for each strain 
CENPK7D <- read_csv("~/Data/CENPK7D/CENPK7D_phenotype.csv") 
ETRED <- read_csv("~/Data/Ethanol Red/ETRED_phenotype.csv") 
ETRED[is.na(ETRED)] <- 0 #correcting for NA data 
PE2 <- read_csv("~/Data/PE-2/PE2_phenotype.csv") 

#creating a single table with all the values 
strain_phenomics_data <- CENPK7D %>% 
  add_row(ETRED) %>%
  add_row(PE2) %>%
  rename(µmax=mumax) #correcting variable names 

```

## Studying ouliers 

```{r, results='hide'}
#study on outliers (example with CDW), visualization with boxplot
out_CDW <-boxplot.stats(strain_phenomics_data$CDW)$out
boxplot(strain_phenomics_data$CDW,
        ylab = "CDW")
mtext(paste("Outliers: ", paste(out_CDW, collapse = ", ")))

# checking the outliers within certain strains and phenotype with quantile method. Comparing the outlier values with the mean.
#cell dry weight 
CDW_out <- which(strain_phenomics_data$CDW < quantile(strain_phenomics_data$CDW, 0.001) | strain_phenomics_data$CDW > quantile(strain_phenomics_data$CDW, 0.999))
strain_phenomics_data[CDW_out,]
mean(strain_phenomics_data$CDW)
# maximum specific growth rate
mumax_out <- which(strain_phenomics_data$µmax < quantile(strain_phenomics_data$µmax, 0.001) | strain_phenomics_data$µmax > quantile(strain_phenomics_data$µmax, 0.999))
strain_phenomics_data[mumax_out,]
mean(strain_phenomics_data$µmax)
# lag phase
lag_out <- which(strain_phenomics_data$lag < quantile(strain_phenomics_data$lag, 0.001) | strain_phenomics_data$lag > quantile(strain_phenomics_data$lag, 0.999))
strain_phenomics_data[lag_out,]
mean(strain_phenomics_data$lag)
# ethanol yield
Yp_out <- which(strain_phenomics_data$Yp < quantile(strain_phenomics_data$Yp, 0.001, na.rm = TRUE) | strain_phenomics_data$Yp > quantile(strain_phenomics_data$Yp, 0.999, na.rm = TRUE))
strain_phenomics_data[Yp_out,]
mean(strain_phenomics_data$Yp)
#biomass yield
Yx_out <- which(strain_phenomics_data$Yx < quantile(strain_phenomics_data$Yx, 0.001) | strain_phenomics_data$Yx > quantile(strain_phenomics_data$Yx, 0.999))
strain_phenomics_data[Yx_out,]
mean(strain_phenomics_data$Yx)

#correcting lag phase 0 values with NAs
strain_phenomics_data$lag[strain_phenomics_data$lag == 0] <- NA
```

## Analysing the strain performance 

```{r, results='hide'}
#Plotting the summary statistics of the variables 
phenotype_reshaped <- gather(strain_phenomics_data, key="variable", value="value", c("µmax", "lag", "Yx", "Yp", "CDW"))

ggplot(phenotype_reshaped, aes(x=strain, y=value, fill=strain)) +
  geom_boxplot()+
  facet_wrap(~variable,  nrow=5, strip.position = "left", scales ="free_y") +
  scale_fill_manual(values=c("#00AFBB", "#E7B800", "#FC4E07"))+
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )
#plotting variable distribution across the conditions
theme_set(theme_ridges())

ggplot(phenotype_reshaped, aes(x = value, y = strain)) +
  geom_density_ridges(aes(fill = strain), jittered_points = TRUE, position = position_points_jitter(height = 0),
                      point_shape = '|', point_size = 3, 
                      point_alpha = 1, alpha = 0.7) +
  scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07")) +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )+
  facet_wrap(~variable,  ncol=1, nrow = 5, strip.position = "left", scales = "free_x") 

#plotting phenotype across conditions
phenotype.means <- aggregate(phenotype_reshaped,
                             by = list(phenotype_reshaped$variable, phenotype_reshaped$strain),
                             FUN = mean) %>%
  select(-variable,-strain) %>%
  rename(variable=Group.1)%>%
  rename(strain=Group.2)

ggdotchart(phenotype_reshaped, x = "condition", y = "value",
           color = "group",                                # Color by groups
           palette = c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"), # Custom color palette
           add = "segments",                             # Add segments from y = 0 to dots
           add.params = list(color = "lightgray", size = 2), # Change segment color and size
           group = "group",                                # Order by groups
           dot.size = 6,                                 # Large dot size
           ggtheme = theme_light()                        # ggplot2 theme
)+
  geom_hline(yintercept = 0, linetype = 2, color = "lightgray") +
  facet_grid(vars(variable), vars(strain), scales = "free_y")+
  geom_hline(data = phenotype.means, aes(yintercept = value),
             linetype = 2, size=0.9)+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )

#plotting the control condition and the relative performance values as similar boxplot chart
ggplot(phenotype_reshaped, aes(x=strain, y=value))+
  geom_violin(alpha=0.5, color="gray")+
  geom_jitter(alpha=0.8, aes(color=group),
              position = position_jitter(width = 0.1))+
  scale_color_manual(values=c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"))+
  facet_wrap(vars(variable), scales = "free_y") +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )+
  stat_summary(fun = "mean",
               geom = "crossbar", 
               width = 0.5,
               linetype=2,
               colour = "black")+
  geom_signif(
    comparisons = list(c("ETRED", "CENPK7D"),c("PE2","ETRED"), c("CENPK7D", "PE2")),
    test="wilcox.test",
    map_signif_level = TRUE,
    textsize=5
  )
```

## Visualization of the normalized mean phenotype with a heatmap

```{r, results='hide'}
#function to normalize the data between 0 and 1 
normalize01 <- function(x){
  (x-min(x))/(max(x)-min(x))
}
normalize <- function(data){
  normalized_robustness <- data[,c(1:9)] %>%
    mutate(CDW_n=normalize01(CDW)) %>%
    mutate(µmax_n=normalize01(mumax)) %>%
    mutate(lag_n=normalize01(lag)) %>%
    mutate(Yx_n=normalize01(Yx)) %>%
    mutate(Yp_n = ifelse(is.na(Yp), 0, Yp)) %>%
    mutate(Yp_n=normalize01(Yp))
}

#applying normalization on the different strains 
norm_CENPK7D <- normalize(CENPK7D)
norm_ETRED <- normalize(ETRED)
norm_PE2 <- normalize(PE2)
#creating one table with all the values
norm01 <- norm_CENPK7D %>% 
  add_row(norm_ETRED) %>%
  add_row(norm_PE2) 

norm_reshaped <- gather(norm01, key="variable", value="value", c("µmax_n", "lag_n", "Yx_n", "Yp_n", "CDW_n"))

#visualization of the normalized phenotype through heatmap
ggplot(data = norm_reshaped,
       mapping = aes(x = strain,
                     y = condition,
                     fill = (value))) +
  geom_tile() +
  xlab(label = "Sample") +
  facet_grid( ~ variable) +
  scale_fill_distiller(palette = "RdYlBu")+
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )

```

## Quantification of Robustness with the formula: R=-F=-sd^2/mean

```{r, results='hide'}
#calculating the mean and the standard deviation for each variable, strain and replicate 
R_m_sd <- strain_phenomics_data %>%
  gather(key="variable", value="value", c("µmax", "lag", "Yx", "Yp", "CDW")) %>%
  group_by(strain, variable, replicate) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) %>%
  mutate(stdev=sd(value,  na.rm=TRUE)) %>%
  filter(condition == "Glu_65" ) %>%
  select(-c(condition, group, strain_type, value))

#calculating R=-F for each replicate. Then, calculating the mean and the standard deviation for each strain and variable
#Each R is normalized for the mean of the values for each variable 
R_Fano <- R_m_sd %>%
  group_by(variable)%>%
  mutate(mean_v=mean(mean, na.rm=TRUE)) %>%
  mutate(robustness=-stdev^2/mean, na.rm=TRUE)%>%
  mutate(robustness_n=robustness/mean_v, na.rm=TRUE) %>%
  ungroup()%>%
  group_by(strain, variable) %>%
  mutate(mean_R=mean(robustness_n, na.rm=TRUE)) %>%
  mutate(sem_R=sd(robustness_n)/sqrt(3),na.rm=TRUE) 

# Calculating significant difference between the strains 
p_µmax <- R_Fano %>% 
  filter(variable == "µmax") 
pairwise.t.test(p_µmax$robustness_n, p_µmax$strain, p.adjust.method = "bonferroni")

p_CDW <- R_Fano %>% 
  filter(variable == "CDW") 
pairwise.t.test(p_CDW$robustness_n, p_CDW$strain)

p_lag <- R_Fano %>% 
  filter(variable == "lag") 
pairwise.t.test(p_lag$robustness_n, p_lag$strain, na.rm=TRUE)

p_Yx <- R_Fano %>% 
  filter(variable == "Yx") 
pairwise.t.test(p_Yx$robustness_n, p_Yx$strain)

p_Yp <- R_Fano %>% 
  filter(variable == "Yp") 
pairwise.t.test(p_Yp$robustness_n, p_Yp$strain)

#plotting Robustness
R_Fano <- R_Fano %>%
  filter(replicate==1) 

ggplot(R_Fano, fill=strain) +
  geom_bar(aes(x=strain, y=mean_R, fill=strain), stat="identity") +
  geom_errorbar(aes(x=strain, ymin=mean_R-sem_R, ymax=mean_R+sem_R), width=0.05, colour="red", alpha=0.9, size=1.3)+
  facet_grid(col = vars(variable), scales = "free_y") +
  scale_fill_manual(values = c('#bdbdbd', '#636363', '#303030')) +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )

```


## Quantification of robusntess as R = 1-coefficient of variation = 1-sd/mean

```{r, results='hide'}
#calculating R=-F for each replicate. Then, calculating the mean and the standard deviation for each strain and variable
R_CV <- R_m_sd %>%
  dplyr::mutate(robustness=1 - stdev/mean) %>%
  group_by(strain, variable) %>%
  mutate(mean_RCV=mean(robustness, na.rm=TRUE)) %>%
  mutate(sd_RCV=sd(robustness, na.rm=TRUE)) %>%
  filter(replicate==1)

#plotting Robustness
ggplot(R_CV, fill=strain) +
  geom_bar(aes(x=strain, y=mean_RCV, fill=strain), stat="identity") +
  geom_errorbar(aes(x=strain, ymin=mean_RCV-sd_RCV, ymax=mean_RCV+sd_RCV), width=0.05, colour="red", alpha=0.9, size=1.3)+
  facet_grid(rows = vars(variable), scales = "free_y") +
  scale_fill_manual(values = c('#bdbdbd', '#636363', '#303030')) +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )
```

## Robusntess quantification with Kitano's formula 

```{r, results='hide'}
#importing the data
CENPK7D_K <- CENPK7D %>%
  mutate(D_CDW = (CDW / mean(CDW[condition == "Glu_20"]))) %>%
  mutate(D_Yx = (Yx / mean(Yx[condition == "Glu_20"]))) %>%
  mutate(D_µmax = (mumax / mean(mumax[condition == "Glu_20"]))) %>%
  mutate(D_Yp = (Yp / mean(Yp[condition == "Glu_20"]))) %>%
  mutate(D_lag = (lag / mean(lag[condition == "Glu_20"])))
ETRED_K <- ETRED %>%
  mutate(D_CDW = (CDW / mean(CDW[condition == "Glu_20"]))) %>%
  mutate(D_Yx = (Yx / mean(Yx[condition == "Glu_20"]))) %>%
  mutate(D_µmax = (mumax / mean(mumax[condition == "Glu_20"]))) %>%
  mutate(D_Yp = (Yp / mean(Yp[condition == "Glu_20"]))) %>%
  mutate(D_lag = (lag / mean(lag[condition == "Glu_20"])))
PE2_K <- PE2 %>%
  mutate(D_CDW = (CDW / mean(CDW[condition == "Glu_20"]))) %>%
  mutate(D_Yx = (Yx / mean(Yx[condition == "Glu_20"]))) %>%
  mutate(D_µmax = (mumax / mean(mumax[condition == "Glu_20"]))) %>%
  mutate(D_Yp = (Yp / mean(Yp[condition == "Glu_20"]))) %>%
  mutate(D_lag = (lag / mean(lag[condition == "Glu_20"])))

#creating a single table
strain_phenomics_data_K <- CENPK7D_K %>%
  add_row(ETRED_K) %>%
  add_row(PE2_K) %>%
  rename(µmax=mumax)
#calculating the sum of the ratios for each strain, replicate and variable 
#Then, calculating the mean and the standard deviation of each of the sum values
R_K <- strain_phenomics_data_K %>%
  gather(key="variable", value="value", c("D_µmax", "D_lag", "D_Yx", "D_Yp", "D_CDW")) %>%
  group_by(strain, variable, replicate) %>%
  mutate(sum=sum(value))%>%
  filter(condition == "Glu_65" ) %>%
  select(-c(condition, group, strain_type, value, CDW,Yx,Yp,µmax,lag))%>%
  group_by(strain, variable) %>%
  mutate(mean_RK=mean(sum)) %>%
  mutate(sd_RK=sd(sum)) %>%
  filter(replicate==1)

#plotting 
ggplot(R_K, fill=strain) +
  geom_bar(aes(x=strain, y=mean_RK, fill=strain), stat="identity") +
  geom_errorbar(aes(x=strain, ymin=mean_RK-sd_RK, ymax=mean_RK+sd_RK), width=0.05, colour="red", alpha=0.9, size=1.3)+
  facet_grid(rows = vars(variable), scales = "free_y") +
  scale_fill_manual(values = c('#bdbdbd', '#636363', '#303030')) +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )

#correcting the calculations above with the mean instead of the sum
R_Km <- strain_phenomics_data_K %>%
  gather(key="variable", value="value", c("D_µmax", "D_lag", "D_Yx", "D_Yp", "D_CDW")) %>%
  group_by(strain, variable, replicate) %>%
  mutate(meanK=mean(value))%>%
  filter(condition == "Glu_65" ) %>%
  select(-c(condition, group, strain_type, value, CDW,Yx,Yp,µmax,lag))%>%
  group_by(strain, variable) %>%
  mutate(mean_RKm=mean(meanK)) %>%
  mutate(sd_RKm=sd(meanK)) %>%
  filter(replicate==1)

#plotting 
ggplot(R_Km, fill=strain) +
  geom_bar(aes(x=strain, y=mean_RKm, fill=strain), stat="identity") +
  geom_errorbar(aes(x=strain, ymin=mean_RKm-sd_RKm, ymax=mean_RKm+sd_RKm), width=0.05, colour="red", alpha=0.9, size=1.3)+
  facet_grid(rows = vars(variable), scales = "free_y") +
  scale_fill_manual(values = c('#bdbdbd', '#636363', '#303030')) +
  theme_light()+
  theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 10, color = "black", face = "bold"),
    strip.text.y = element_text(size = 10, color = "black", face = "bold")
  )
```

